---
title: "From monomers to mechanisms: an end-to-end ABC workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From monomers to mechanisms: an end-to-end ABC workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  out.width  = "100%",
  dpi       = 150
)
set.seed(20260817)
library(data.table)
library(reCYCLing)
has_bio   <- requireNamespace("Biostrings", quietly = TRUE)
has_abcrf <- requireNamespace("abcrf", quietly = TRUE)
```

## The question, and the workflow

Large tandem repeat arrays — centromeric satellites, knobs, subtelomeric
repeats — are built by mechanisms nobody can watch directly: tandem
duplication of multi-monomer units, longer-range duplication within a
chromosome, and exchange of blocks between chromosomes. What we *can*
observe is the end product: thousands to millions of monomers whose
sequences and positions carry a fossil record of those events.

reCYCLing turns that record into inference by simulation. The workflow has
four steps, and this vignette runs all of them end to end at toy scale:

1. **Observe** — extract per-monomer sequences from real assemblies (e.g.
   with [TRASH](https://github.com/vlothec/TRASH)) and import them
   (`assemble_genome_dt()`).
2. **Describe** — reduce a genome of monomers to a fixed-length vector of
   summary statistics (`repeat_genome_stats()`, `repeat_abc_predictors()`).
   The *same* functions score real and simulated genomes, so the two are
   comparable by construction.
3. **Simulate** — draw mechanism parameters from priors, evolve simulated
   genomes (`sim_genome()`), and score them, building a reference table.
4. **Infer** — use random-forest ABC (the `abcrf` package) for model choice
   ("is mechanism X present at all?") and parameter posteriors, then check
   the fit by simulating at the posterior estimates (posterior-predictive
   check).

Everything below runs in a couple of minutes on a laptop. The final section
shows how the same code scales to real inference (tens of thousands of
simulations on a cluster).

## Step 0 — the generative model

`sim_genome()` evolves a multi-chromosome genome of monomers forward in
time. Each chromosome starts as a small array of identical copies of one
ancestral monomer and grows toward a target size under:

| Mechanism | Rate parameter | Size parameter |
|---|---|---|
| Local (tandem) duplication of a multi-monomer unit | `p_local_dup` (per monomer per generation) | `local_dist` |
| Long-range duplication within a chromosome | `p_distal_dup` (per chromosome per generation) | `distal_dist` |
| Inter-chromosomal exchange (copy a block to another chromosome) | `p_translocation` (per genome per generation) | `transloc_chunk_mean` |
| Point mutation (the molecular clock) | `mu_total` (per base per generation) | — |

A size governor (deletion) keeps each chromosome near its target once
reached, so arrays keep *turning over* rather than growing without bound —
the regime real satellite arrays live in.

The output is the package's central **data contract**: a `data.table` with
one row per monomer and columns `chrom` (integer 1..K), `num` (position
within the chromosome), `bponly` (gapless sequence), and `dir` (strand).

```{r sim_demo}
g <- sim_genome(
  K             = 3,
  target_sizes  = c(600, 450, 300),
  n_generations = 250,
  mu_total      = 5e-5,
  p_local_dup   = 2.5e-4,
  local_dist    = list(type = "gamma", shape = 2, scale = 8 / 2),
  p_distal_dup  = 0.05,
  distal_dist   = list(type = "gamma", shape = 2, scale = 300 / 2),
  p_translocation     = 0.2,
  transloc_chunk_mean = 100
)
head(g, 3)
table(g$chrom)
```

## Step 1 — observed data

### Real assemblies

For real genomes, run a monomer annotator such as TRASH on each
chromosome-scale assembly to get one fasta of monomers per chromosome
(optionally aligned, i.e. gapped). `assemble_genome_dt()` parses the fasta
headers — split on `sep`, with the chromosome id, monomer position, and
strand at fixed field positions — and returns the data contract above:

```{r import_real, eval=FALSE}
files <- list.files("monomers/", pattern = "\\.fasta$", full.names = TRUE)
obs <- assemble_genome_dt(files,
                          chrom_field = 3,   # e.g. "sp_hap1_Chr01_00042_..._+"
                          num_field   = 4,
                          dir_field   = 7,
                          sep         = "_")
```

For a single array (one chromosome) there is a one-call shortcut,
`repeat_stats_from_fasta()`, which goes straight from fasta to the stat
object of Step 2.

### A round trip you can run

To show the import path without shipping data, we write one simulated
chromosome out as a TRASH-style monomer fasta and read it back:

```{r roundtrip, eval=has_bio}
fa <- tempfile(fileext = ".fasta")
sub <- g[chrom == 1L]
writeLines(paste0(">mono_demo_Chr01_", sub$num, "_x_x_", sub$dir, "\n", sub$bponly), fa)
back <- assemble_genome_dt(fa, chrom_field = 3, num_field = 4, dir_field = 7)
identical(back$bponly, sub$bponly)
```

### Pseudo-observed genomes with known truth

For the rest of this vignette we use two simulated genomes as
"observations", so we know the truth the inference should recover:
one evolved **with** inter-chromosomal exchange, one **without**.

```{r pseudo_obs}
make_obs <- function(p_xch, seed) {
  set.seed(seed)
  sim_genome(K = 3, target_sizes = c(600, 450, 300), n_generations = 250,
             mu_total = 5e-5, p_local_dup = 2.5e-4,
             local_dist  = list(type = "gamma", shape = 2, scale = 8 / 2),
             p_distal_dup = 0.05,
             distal_dist = list(type = "gamma", shape = 2, scale = 300 / 2),
             p_translocation = p_xch, transloc_chunk_mean = 100)
}
obs_with    <- make_obs(p_xch = 0.2, seed = 101)  # exchange PRESENT
obs_without <- make_obs(p_xch = 0,   seed = 102)  # exchange ABSENT
```

## Step 2 — summary statistics

`repeat_genome_stats()` reduces a genome to a stat object; the two
flatteners turn it into named numeric vectors:

* `repeat_genome_predictors()` — the **full catalog** (every statistic the
  package computes, fixed length, NA-safe). Use this for reference tables:
  storing everything means you can change the fitted subset later without
  re-simulating, and unfitted statistics become free posterior-predictive
  checks.
* `repeat_abc_predictors()` — a curated **fit subset** built from three
  interpretable families, designed for mechanism inference:

| Family | Statistics | What it measures |
|---|---|---|
| Divergence (clock) | `gdv01`–`gdv11`, `mean_div_bp_m` | distribution of monomer divergence from consensus — array age / turnover |
| Within-chromosome pair distance | `wxd01`–`wxd20` | log-binned distances between identical monomers — local vs long-range duplication |
| Cross-chromosome sharing | `frac_pairs_xchrom`, `frac_genome_xchrom`, `log_n_xchrom_fam`, `log_xchrom_jacc_mean` | identical monomers shared between chromosomes — exchange |
| Size context | `log_array_size_m`, `log_n_chrom` | scale of the genome |

```{r stats_demo}
st <- repeat_genome_stats(obs_with)
obs_stats <- rbind(
  with_exchange    = repeat_abc_predictors(st),
  without_exchange = repeat_abc_predictors(repeat_genome_stats(obs_without))
)
length(repeat_genome_predictors(st))   # full catalog
ncol(obs_stats)                        # fit subset
round(obs_stats[, c("mean_div_bp_m", "frac_pairs_xchrom", "log_array_size_m")], 3)
```

The two genomes have the same age and size, but the exchange signal
(`frac_pairs_xchrom`) separates them already. The inference below
formalizes that.

## Step 3 — a reference table

Draw parameters from priors, simulate, score. Two design points worth
copying at any scale:

* **Spike and slab.** Each optional mechanism is switched *off* entirely in
  a fraction of simulations (here 1/3). That turns "is exchange present?"
  into a model-choice problem a classification forest can answer.
* **Store the full catalog** (`repeat_genome_predictors()`), select the fit
  subset (`repeat_abc_predictors()`) downstream.

```{r reftable, cache=FALSE}
rlunif <- function(n, lo, hi) 10^runif(n, log10(lo), log10(hi))
SPIKE <- 1/3

draw_prior <- function() {
  has_xch <- runif(1) >= SPIKE
  list(n_generations = round(runif(1, 80, 400)),
       local_mean    = rlunif(1, 2, 50),
       p_distal      = rlunif(1, 1e-3, 0.3),
       distal_mean   = rlunif(1, 50, 1000),
       p_xch         = if (has_xch) rlunif(1, 1e-2, 0.9) else 0,
       xch_mean      = if (has_xch) rlunif(1, 10, 500) else NA_real_)
}

simulate_one <- function(pr) {
  sim <- sim_genome(K = 3, target_sizes = round(rlunif(3, 200, 800)),
                    n_generations = pr$n_generations,
                    mu_total = 5e-5, p_local_dup = 2.5e-4,
                    local_dist  = list(type = "gamma", shape = 2, scale = pr$local_mean / 2),
                    p_distal_dup = pr$p_distal,
                    distal_dist = list(type = "gamma", shape = 2, scale = pr$distal_mean / 2),
                    p_translocation     = pr$p_xch,
                    transloc_chunk_mean = if (is.na(pr$xch_mean)) 100 else pr$xch_mean)
  repeat_abc_predictors(repeat_genome_stats(sim))
}

N <- 240
t0 <- Sys.time()
rows <- lapply(seq_len(N), function(i) {
  pr <- draw_prior()
  pred <- tryCatch(simulate_one(pr), error = function(e) NULL)
  if (is.null(pred)) return(NULL)
  c(unlist(pr[c("n_generations", "local_mean", "p_xch")]), pred)
})
ref <- as.data.table(do.call(rbind, Filter(Negate(is.null), rows)))
ref <- ref[complete.cases(ref)]
cat(sprintf("%d usable simulations in %.1f min\n",
            nrow(ref), as.numeric(Sys.time() - t0, units = "mins")))

# statistics that never vary at this toy scale (fixed K, empty divergence
# bins) carry no information here — drop them before training
stat_cols <- setdiff(colnames(ref), c("n_generations", "local_mean", "p_xch"))
stat_cols <- stat_cols[sapply(ref[, ..stat_cols], sd) > 0]
```

## Step 4 — inference with random forests

Random-forest ABC (Pudlo et al. 2016; Raynal et al. 2019, both via the
`abcrf` package) uses the reference table as a training set: a
classification forest votes on mechanism presence, a regression forest
estimates parameter posteriors. Forests tolerate many correlated
statistics, so no statistic pre-selection is needed.

### Model choice: is exchange present?

```{r model_choice, eval=has_abcrf}
library(abcrf)
train <- data.frame(mod = factor(ifelse(ref$p_xch > 0, "present", "absent")),
                    ref[, ..stat_cols])
obs_df <- data.frame(obs_stats)[, stat_cols]

rf_mc  <- abcrf(mod ~ ., data = train, ntree = 200, paral = FALSE)
votes  <- predict(rf_mc, obs_df, train, ntree = 200, paral = FALSE)
data.frame(genome   = rownames(obs_stats),
           truth    = c("present", "absent"),
           P_present = round(votes$vote[, "present"] / rowSums(votes$vote), 2))
```

The forest assigns a high probability of exchange to the genome simulated
with it and a low one to the genome without — with only
`r nrow(ref)` training simulations.

### Parameter posteriors

A regression forest per parameter gives posterior medians and credible
intervals. Here: array age (`n_generations`) and the local duplication
unit size (`local_mean`), with the known truth alongside.

```{r regression, eval=has_abcrf}
posterior <- function(param, obs_row) {
  tr <- data.frame(r = ref[[param]], ref[, ..stat_cols])
  rf <- regAbcrf(r ~ ., data = tr, ntree = 200, paral = FALSE)
  pr <- predict(rf, obs_df[obs_row, , drop = FALSE], tr,
                quantiles = c(0.025, 0.5, 0.975), paral = FALSE)
  c(median = pr$med[1], lo = pr$quantiles[1], hi = pr$quantiles[3])
}
res <- rbind(
  age_generations = posterior("n_generations", 1),
  local_mean      = posterior("local_mean", 1)
)
cbind(round(res, 1), truth = c(250, 8))
```

Even this toy table recovers age well (the divergence family is a clock)
and puts the local unit size in the right range; at real scale (30k
simulations) both tighten substantially, and cross-validation on held-out
simulations quantifies exactly how much (see `abcrf::err.regAbcrf` and the
scaling section below).

## Step 5 — posterior-predictive check

The loop closes by simulating at the inferred values and asking whether the
result resembles the observation — including in statistics that were
*never fitted*. Here we overlay the fitted within-chromosome distance
histogram (`wxd`).

```{r postpred, eval=has_abcrf, fig.height=3.5}
library(ggplot2)
sim_pp <- sim_genome(K = 3, target_sizes = c(600, 450, 300),
                     n_generations = round(res["age_generations", "median"]),
                     mu_total = 5e-5, p_local_dup = 2.5e-4,
                     local_dist  = list(type = "gamma", shape = 2,
                                        scale = res["local_mean", "median"] / 2),
                     p_distal_dup = 0.05,
                     distal_dist = list(type = "gamma", shape = 2, scale = 300 / 2),
                     p_translocation = 0.2, transloc_chunk_mean = 100)
pp <- repeat_abc_predictors(repeat_genome_stats(sim_pp))

wxd <- grep("^wxd", stat_cols, value = TRUE)
dt <- rbind(data.table(bin = seq_along(wxd), p = obs_stats["with_exchange", wxd], who = "observed"),
            data.table(bin = seq_along(wxd), p = pp[wxd], who = "posterior-predictive"))
ggplot(dt, aes(bin, p, fill = who)) +
  geom_col(position = "dodge") +
  labs(x = "log10 identical-pair distance (binned)", y = "proportion of pairs",
       fill = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

A systematic mismatch — especially in held-out statistics — is not a
failure of the workflow but its most useful output: it points at a
mechanism the model lacks.

## Scaling to real inference

The toy above is the real pipeline in miniature. At publication scale the
same calls change only in size and placement:

```{r scale, eval=FALSE}
## Reference table: ~30,000 simulations, full catalog, on a cluster.
## One genome per row; checkpoint batches to .rds so preempted jobs resume.
pred <- repeat_genome_predictors(repeat_genome_stats(sim))   # store EVERYTHING

## Real data: one fit per genome/haplotype.
obs  <- assemble_genome_dt(chromosome_fastas)                # TRASH monomers
v    <- repeat_abc_predictors(repeat_genome_stats(obs))

## Inference: abcrf with ntree = 500-1000, parallel;
## validate with ~1000 held-out simulations as pseudo-observations
## (parameter recovery, CI coverage, false-positive rate of model choice).
```

Practices that matter at scale:

* **Fit per genome, then pool.** With several genomes (haplotypes,
  accessions, species), fit each independently and compare or pool
  posteriors — independent haplotypes of the same genotype are a built-in
  replication check.
* **Hold statistics out.** Fit on the three families above; everything else
  in the full catalog scores the fit afterwards.
* **Validate the estimator.** Treat held-out simulations as
  pseudo-observations: parameter recovery, credible-interval coverage, and
  — for model choice — how often a truly-absent mechanism reaches the vote
  level of your real genomes.

## Function reference

| Step | Function | Purpose |
|---|---|---|
| Simulate | `sim_genome()` | multi-chromosome genome under the mechanism model |
| Simulate | `run_sim_ps()` | single array, with identical-pair tracking and plots (see the intro vignette) |
| Import | `assemble_genome_dt()` | monomer fastas (TRASH-style headers) → genome data contract |
| Import | `repeat_stats_from_fasta()` | single-array fasta → stat object, one call |
| Describe | `repeat_genome_stats()` | genome → stat object |
| Describe | `repeat_genome_predictors()` | stat object → full catalog vector |
| Describe | `repeat_abc_predictors()` | stat object → curated fit subset |
| Describe | `repeat_summary_stats()`, `repeat_predictors()` | array-level equivalents |
| Visualize | `plot_repeat_fingerprint()`, `plot_repeat_barcode()`, `plot_ps_summary()` | array structure plots |
