---
title: "Equilibrium dynamics in tandem repeat arrays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Equilibrium dynamics in tandem repeat arrays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 4,
  out.width  = "100%",
  dpi = 150
)
set.seed(42)
library(data.table)
library(ggplot2)
library(reCYCLing)
```

## The growth problem

In the standard reCYCLing simulation, a tandem repeat array grows from a small
initial copy number until it reaches a target size (`max_units`). But in real
genomes, repeat arrays don't grow without bound — they reach a characteristic
size that reflects a balance between expansion and contraction forces.

Can we reproduce this equilibrium in simulation?

## Two kinds of duplication, one kind of deletion

The simulator has three forces that change array size:

| Force | Rate type | Effect on size |
|-------|-----------|----------------|
| **Local duplication** | Per-unit per-gen | Grows faster as array gets bigger |
| **Distal duplication** | Per-array per-gen | Constant injection regardless of size |
| **Chunk deletion** | Per-unit per-gen | Shrinks faster as array gets bigger |

Local duplication and deletion both scale with array size. If their rates and
chunk sizes are similar, they roughly cancel — but the array does a random walk
that eventually hits zero (extinction).

**Distal duplication is the key to equilibrium.** Because it's a per-array event
(one crossover per generation, independent of array size), it acts as a constant
source of new material. As the array grows, per-unit deletion increases until it
balances the distal input. The result: a stable equilibrium size.

## Demonstrating equilibrium

Let's run simulations with different deletion rates while keeping distal
duplication fixed, and watch the array size over time.

```{r equilibrium_sims, cache=TRUE}
params <- list(
  list(name = "Low deletion (5e-6)",    pdel = 5e-6),
  list(name = "Medium deletion (5e-5)", pdel = 5e-5),
  list(name = "High deletion (7e-5)",   pdel = 7e-5),
  list(name = "Very high deletion (2e-4)", pdel = 2e-4)
)

trajectories <- rbindlist(lapply(params, function(p) {
  sim <- suppressWarnings(run_sim_ps(
    max_units = 200000, max_t = 5000, hard_cap = 200000,
    init_l = 178, init_k0 = 10,
    p_local_dup = 1e-5,      # low local dup
    p_distal_dup = 0.05,     # 5% chance of crossover per generation
    p_del_chunk = p$pdel,
    mu_total = 5e-5,
    verbose = FALSE, compute_pairs = FALSE
  ))
  data.table(
    gen = seq_along(sim$L_vec),
    size = sim$L_vec,
    scenario = p$name
  )
}))
```

```{r plot_trajectories, fig.height=5}
ggplot(trajectories, aes(x = gen, y = size, colour = scenario)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_y_log10(labels = scales::comma) +
  theme_classic(base_size = 12) +
  labs(x = "Generation", y = "Array size (monomers)",
       colour = "Deletion rate",
       title = "Array size trajectories under different deletion rates",
       subtitle = "Distal dup = 0.05/gen, local dup = 1e-5/unit/gen") +
  theme(legend.position = "bottom")
```

With low deletion, the array grows without bound. With higher deletion rates,
the array reaches an equilibrium size where distal duplication input balances
deletion loss. The equilibrium size decreases as deletion rate increases.

## Equilibrium size depends on the balance of forces

Let's measure the equilibrium size across a range of deletion rates.

```{r eq_sweep, cache=TRUE}
del_rates <- 10^seq(-5.5, -3, by = 0.25)

eq_data <- rbindlist(lapply(del_rates, function(pdel) {
  sim <- suppressWarnings(run_sim_ps(
    max_units = 200000, max_t = 5000, hard_cap = 200000,
    init_l = 178, init_k0 = 10,
    p_local_dup = 1e-5, p_distal_dup = 0.05,
    p_del_chunk = pdel, mu_total = 5e-5,
    verbose = FALSE, compute_pairs = FALSE
  ))
  lv <- sim$L_vec
  n <- length(lv)
  # Use last 20% as equilibrium estimate
  tail_start <- max(1, as.integer(n * 0.8))
  tail_vals <- lv[tail_start:n]
  data.table(
    p_del = pdel,
    eq_size = mean(tail_vals),
    eq_sd = sd(tail_vals),
    n_gens = n,
    final = lv[n]
  )
}))
```

```{r plot_eq_size}
ggplot(eq_data, aes(x = p_del, y = eq_size)) +
  geom_pointrange(aes(ymin = eq_size - eq_sd, ymax = eq_size + eq_sd)) +
  scale_x_log10() +
  scale_y_log10(labels = scales::comma) +
  theme_classic(base_size = 12) +
  labs(x = "Deletion rate (per unit per gen)",
       y = "Equilibrium array size",
       title = "Equilibrium size vs deletion rate",
       subtitle = "Error bars = SD of size in last 20% of trajectory")
```

The relationship is roughly inverse: doubling the deletion rate halves the
equilibrium size. This makes intuitive sense — the equilibrium occurs where
total deletion rate (per-unit rate × array size) equals the distal duplication
input.

## Tracking array properties over time

To understand how the array evolves toward equilibrium, we can track multiple
properties over time: array size, mutation load, and monomer diversity. We'll
sample these at regular checkpoints by running the simulation in stages.

```{r time_series, cache=TRUE}
library(cowplot)

# Run equilibrium sim with checkpoints every 100 gens
# Lower mutation rate for more realistic load values
max_gen <- 3000
checkpoint_every <- 100
checkpoints <- seq(checkpoint_every, max_gen, by = checkpoint_every)

# We'll run the full sim once, then use L_vec for size trajectory.
# For load and diversity, we need to run in stages using the R backend.
sim_eq <- suppressWarnings(run_sim_ps(
  max_units = 200000, max_t = max_gen, hard_cap = 200000,
  init_l = 178, init_k0 = 10,
  p_local_dup = 1e-5, p_distal_dup = 0.05,
  p_del_chunk = 7e-5, mu_total = 1e-5,
  verbose = FALSE, compute_pairs = FALSE
))

# Size trajectory from L_vec
traj_dt <- data.table(gen = seq_along(sim_eq$L_vec), size = sim_eq$L_vec)

# For the final state, compute load and copy number
mono <- sim_eq$monomers
ct <- counts_long_nogap(mono$bponly)
cons <- ct[symbol == consensus][order(pos)]
load <- mu_load_from_consensus(mono$bponly, cons)

cat("Final array size:", nrow(mono), "\n")
cat("Generations:", length(sim_eq$L_vec), "\n")
cat("Mean mutation load:", round(mean(load), 1), "\n")
cat("Unique sequences:", uniqueN(mono$bponly), "\n")

# Copy number distribution
cn <- mono[, .N, by = bponly][order(-N)]
cat("Mean copy number:", round(mean(cn$N), 1), "\n")
cat("Max copy number:", max(cn$N), "\n")
```

```{r plot_trajectory, fig.height=3}
ggplot(traj_dt, aes(x = gen, y = size)) +
  geom_line(colour = "steelblue", linewidth = 0.5) +
  theme_classic(base_size = 11) +
  labs(x = "Generation", y = "Array size (monomers)",
       title = "Array size over time — approach to equilibrium")
```

The array rapidly expands in the first few hundred generations, then fluctuates
around an equilibrium size determined by the balance of duplication and deletion.

## Mutation load and diversity at the final state

```{r final_state_plots, fig.height=8}
p_load <- ggplot(data.frame(load = load), aes(x = load)) +
  geom_histogram(bins = 40, fill = "firebrick", alpha = 0.7, colour = "white") +
  geom_vline(xintercept = mean(load), linetype = "dashed") +
  theme_classic(base_size = 11) +
  labs(x = "Mismatches to consensus", y = "Count",
       title = paste0("Mutation load (mean = ", round(mean(load), 1), ")"))

p_cn <- ggplot(cn, aes(x = N)) +
  geom_histogram(bins = 50, fill = "darkorange", alpha = 0.7, colour = "white") +
  scale_x_log10() +
  theme_classic(base_size = 11) +
  labs(x = "Copies per unique sequence", y = "Count",
       title = paste0("Copy number distribution (", nrow(cn), " unique sequences)"))

# Shannon entropy as a single number
H <- shannon_entropy(mono$bponly)
p_info <- ggplot() + theme_void() +
  annotate("text", x = 0.5, y = 0.7, size = 5,
           label = paste0("Array size: ", nrow(mono))) +
  annotate("text", x = 0.5, y = 0.5, size = 5,
           label = paste0("Shannon entropy: ", round(H, 2), " bits")) +
  annotate("text", x = 0.5, y = 0.3, size = 5,
           label = paste0("Unique sequences: ", uniqueN(mono$bponly),
                          " / ", nrow(mono)))

plot_grid(p_load, p_cn, p_info, ncol = 1, rel_heights = c(1, 1, 0.5))
```

## Comparing different evolutionary ages

How do array properties change with time? Let's run simulations of different
lengths and compare.

```{r age_comparison, cache=TRUE}
ages <- c(500, 1000, 2000, 3000, 5000)

age_stats <- rbindlist(lapply(ages, function(t_max) {
  sim <- suppressWarnings(run_sim_ps(
    max_units = 200000, max_t = t_max, hard_cap = 200000,
    init_l = 178, init_k0 = 10,
    p_local_dup = 1e-5, p_distal_dup = 0.05,
    p_del_chunk = 7e-5, mu_total = 1e-5,
    verbose = FALSE, compute_pairs = FALSE
  ))
  m <- sim$monomers
  if (nrow(m) < 10) return(NULL)
  ct <- tryCatch(counts_long_nogap(m$bponly), error = function(e) NULL)
  if (is.null(ct)) return(NULL)
  cons <- ct[symbol == consensus][order(pos)]
  ld <- mu_load_from_consensus(m$bponly, cons)
  cn <- m[, .N, by = bponly]
  data.table(
    generations = t_max,
    array_size = nrow(m),
    mean_load = mean(ld),
    median_load = median(ld),
    n_unique = uniqueN(m$bponly),
    mean_copy_number = mean(cn$N),
    max_copy_number = max(cn$N),
    entropy = shannon_entropy(m$bponly)
  )
}))

print(age_stats)
```

```{r plot_age_comparison, fig.height=8}
age_long <- melt(age_stats,
  id.vars = "generations",
  measure.vars = c("array_size", "mean_load", "mean_copy_number", "entropy"))

ggplot(age_long, aes(x = generations, y = value)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  theme_classic(base_size = 11) +
  theme(strip.background = element_blank(), strip.text = element_text(face = "bold")) +
  labs(x = "Generations of evolution", y = NULL,
       title = "Array properties across evolutionary time")
```

Key observations:

- **Array size** rapidly reaches equilibrium (within ~500 gens) and then
  fluctuates around a stable value.
- **Mutation load** increases roughly linearly with time, as expected from
  `mu × init_l × generations`.
- **Mean copy number** may initially increase (duplication homogenises) then
  decrease as mutations diversify.
- **Entropy** increases over time as mutation creates diversity, modulated by
  duplication events that temporarily reduce it.

## The fingerprint at equilibrium

What does the identical-pair structure look like when the array has been
evolving at equilibrium for different lengths of time?

```{r fingerprint_eq, fig.height=8, cache=TRUE}
sim_eq_with_pairs <- sim_eq
sim_eq_with_pairs$ps <- pairs_identical(mono)

plot_ps_summary(sim_eq_with_pairs,
  title = paste0("Equilibrium array (", length(sim_eq$L_vec), " gens)"))
```

## From fossils to simulations: the molecular clock

We can ground the simulation in real evolutionary time using empirical
observations. Here is the logic chain for pistachio knob repeats:

**Known quantities:**

| Observation | Value | Source |
|-------------|-------|--------|
| Ancestor age | ~10 million years | Phylogenetic dating |
| Generation time | ~10 years | Age to reproductive maturity |
| Generations of evolution | ~1,000,000 | 10M yr / 10 yr |
| Monomer length | 178 bp | Sequence data |
| Observed mean mutation load | ~10 mismatches | Empirical from arrays |
| Observed array size | ~20,000 units | Genome assembly |

**Deriving the per-base mutation rate:**

Since `mean_load ≈ mu_per_base × monomer_length × generations`:

```
10 ≈ mu × 178 × 1,000,000
mu ≈ 5.6 × 10^-8 per base per generation
```

This falls squarely within the empirical range for plant per-base mutation
rates (10^-8 to 10^-9), providing independent validation.

**Time compression for simulation:**

We cannot feasibly simulate 1,000,000 generations directly. Instead, we
compress time by scaling the mutation rate up and the generation count down,
preserving the product `mu × t`:

| | Real | Compressed (1000×) |
|-|------|--------------------|
| Mutation rate | 5.6 × 10^-8 | 5.6 × 10^-5 |
| Generations | 1,000,000 | 1,000 |
| mu × t product | 0.056 | 0.056 |
| Expected load (178bp) | 10.0 | 10.0 |

All per-generation rates (duplication, deletion) scale by the same compression
factor, so the *effective* rates in the compressed simulation are 1000× the
real rates.

```{r clock_demo, cache=TRUE}
# Demonstrate: compressed simulation produces the same load
mu_real <- 5.6e-8
compression <- 1000
mu_compressed <- mu_real * compression
t_compressed <- 1000  # = 1M real gens / 1000

sim_clock <- suppressWarnings(run_sim_ps(
  max_units = 200000, max_t = t_compressed, hard_cap = 200000,
  init_l = 178, init_k0 = 10,
  p_local_dup = 1e-5, p_distal_dup = 0.05,
  p_del_chunk = 7e-5, mu_total = mu_compressed,
  verbose = FALSE, compute_pairs = FALSE
))

mono_clock <- sim_clock$monomers
ct_clock <- counts_long_nogap(mono_clock$bponly)
cons_clock <- ct_clock[symbol == consensus][order(pos)]
load_clock <- mu_load_from_consensus(mono_clock$bponly, cons_clock)

cat("Compressed simulation:\n")
cat("  Generations:", length(sim_clock$L_vec), "\n")
cat("  Array size:", nrow(mono_clock), "\n")
cat("  Mean mutation load:", round(mean(load_clock), 1), "\n")
cat("  Expected load (mu × l × t):", round(mu_compressed * 178 * length(sim_clock$L_vec), 1), "\n")
```

## Implications for ABC inference

This molecular clock framework fundamentally changes how we set up the ABC:

**What is fixed (known from data):**

- `mu_total` — derived from mutation rate × compression factor
- `max_t` — derived from ancestor age / generation time / compression factor
- `init_l` — monomer length from sequence data
- Target mutation load — from empirical observation (~10)
- Target array size — from genome assembly (~20,000)

**What the ABC infers:**

- `p_local_dup` — tandem duplication rate (per-unit)
- `p_distal_dup` — crossover/distal duplication rate (per-array)
- `p_del_chunk` — deletion rate (per-unit)
- Chunk-size distribution parameters (shape, scale for each)

**The key insight:** The simulation no longer grows to a target size and stops.
Instead, it runs for a fixed number of generations (representing real
evolutionary time) and the ABC compares the *resulting* array against
observations:

- Does it reach the right size? → constrains deletion/duplication balance
- Does it have the right distance distribution? → constrains chunk sizes
- Does it have the right mutation load? → already locked by the clock
- Does it show bimodal pair distances? → confirms local + distal mechanisms

The equilibrium model means the array size is an emergent property of the
parameter combination, not an input. This is more biologically realistic
and provides stronger parameter identifiability — there is no longer a
degeneracy between "fast rates, few generations" and "slow rates, many
generations" because the number of generations is fixed by the fossil record.
