---
title: "Equilibrium dynamics in tandem repeat arrays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Equilibrium dynamics in tandem repeat arrays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 4,
  out.width  = "100%",
  dpi = 150
)
set.seed(42)
library(data.table)
library(ggplot2)
library(reCYCLing)
```

## The growth problem

In the standard reCYCLing simulation, a tandem repeat array grows from a small
initial copy number until it reaches a target size (`max_units`). But in real
genomes, repeat arrays don't grow without bound — they reach a characteristic
size that reflects a balance between expansion and contraction forces.

Can we reproduce this equilibrium in simulation?

## Two kinds of duplication, one kind of deletion

The simulator has three forces that change array size:

| Force | Rate type | Effect on size |
|-------|-----------|----------------|
| **Local duplication** | Per-unit per-gen | Grows faster as array gets bigger |
| **Distal duplication** | Per-array per-gen | Constant injection regardless of size |
| **Chunk deletion** | Per-unit per-gen | Shrinks faster as array gets bigger |

Local duplication and deletion both scale with array size. If their rates and
chunk sizes are similar, they roughly cancel — but the array does a random walk
that eventually hits zero (extinction).

**Distal duplication is the key to equilibrium.** Because it's a per-array event
(one crossover per generation, independent of array size), it acts as a constant
source of new material. As the array grows, per-unit deletion increases until it
balances the distal input. The result: a stable equilibrium size.

## Demonstrating equilibrium

Let's run simulations with different deletion rates while keeping distal
duplication fixed, and watch the array size over time.

```{r equilibrium_sims, cache=TRUE}
params <- list(
  list(name = "Low deletion (5e-6)",    pdel = 5e-6),
  list(name = "Medium deletion (5e-5)", pdel = 5e-5),
  list(name = "High deletion (7e-5)",   pdel = 7e-5),
  list(name = "Very high deletion (2e-4)", pdel = 2e-4)
)

trajectories <- rbindlist(lapply(params, function(p) {
  sim <- suppressWarnings(run_sim_ps(
    max_units = 200000, max_t = 5000, hard_cap = 200000,
    init_l = 178, init_k0 = 10,
    p_local_dup = 1e-5,      # low local dup
    p_distal_dup = 0.05,     # 5% chance of crossover per generation
    p_del_chunk = p$pdel,
    mu_total = 5e-5,
    verbose = FALSE, compute_pairs = FALSE
  ))
  data.table(
    gen = seq_along(sim$L_vec),
    size = sim$L_vec,
    scenario = p$name
  )
}))
```

```{r plot_trajectories, fig.height=5}
ggplot(trajectories, aes(x = gen, y = size, colour = scenario)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_y_log10(labels = scales::comma) +
  theme_classic(base_size = 12) +
  labs(x = "Generation", y = "Array size (monomers)",
       colour = "Deletion rate",
       title = "Array size trajectories under different deletion rates",
       subtitle = "Distal dup = 0.05/gen, local dup = 1e-5/unit/gen") +
  theme(legend.position = "bottom")
```

With low deletion, the array grows without bound. With higher deletion rates,
the array reaches an equilibrium size where distal duplication input balances
deletion loss. The equilibrium size decreases as deletion rate increases.

## Equilibrium size depends on the balance of forces

Let's measure the equilibrium size across a range of deletion rates.

```{r eq_sweep, cache=TRUE}
del_rates <- 10^seq(-5.5, -3, by = 0.25)

eq_data <- rbindlist(lapply(del_rates, function(pdel) {
  sim <- suppressWarnings(run_sim_ps(
    max_units = 200000, max_t = 5000, hard_cap = 200000,
    init_l = 178, init_k0 = 10,
    p_local_dup = 1e-5, p_distal_dup = 0.05,
    p_del_chunk = pdel, mu_total = 5e-5,
    verbose = FALSE, compute_pairs = FALSE
  ))
  lv <- sim$L_vec
  n <- length(lv)
  # Use last 20% as equilibrium estimate
  tail_start <- max(1, as.integer(n * 0.8))
  tail_vals <- lv[tail_start:n]
  data.table(
    p_del = pdel,
    eq_size = mean(tail_vals),
    eq_sd = sd(tail_vals),
    n_gens = n,
    final = lv[n]
  )
}))
```

```{r plot_eq_size}
ggplot(eq_data, aes(x = p_del, y = eq_size)) +
  geom_pointrange(aes(ymin = eq_size - eq_sd, ymax = eq_size + eq_sd)) +
  scale_x_log10() +
  scale_y_log10(labels = scales::comma) +
  theme_classic(base_size = 12) +
  labs(x = "Deletion rate (per unit per gen)",
       y = "Equilibrium array size",
       title = "Equilibrium size vs deletion rate",
       subtitle = "Error bars = SD of size in last 20% of trajectory")
```

The relationship is roughly inverse: doubling the deletion rate halves the
equilibrium size. This makes intuitive sense — the equilibrium occurs where
total deletion rate (per-unit rate × array size) equals the distal duplication
input.

## Mutation load at equilibrium

At equilibrium, monomers have been accumulating mutations for many generations.
The mutation load reflects the age of the array — how long the evolutionary
process has been running.

```{r eq_load, cache=TRUE}
# Run one long equilibrium simulation and track load over time
sim_eq <- suppressWarnings(run_sim_ps(
  max_units = 200000, max_t = 8000, hard_cap = 200000,
  init_l = 178, init_k0 = 10,
  p_local_dup = 1e-5, p_distal_dup = 0.05,
  p_del_chunk = 7e-5, mu_total = 5e-5,
  verbose = FALSE, compute_pairs = FALSE
))

mono <- sim_eq$monomers
ct <- counts_long_nogap(mono$bponly)
cons <- ct[symbol == consensus][order(pos)]
load <- mu_load_from_consensus(mono$bponly, cons)

cat("Final array size:", nrow(mono), "\n")
cat("Generations:", length(sim_eq$L_vec), "\n")
cat("Mean mutation load:", round(mean(load), 1), "\n")
cat("Median mutation load:", round(median(load), 1), "\n")
```

```{r plot_load_and_trajectory, fig.height=6}
library(cowplot)

p_traj <- ggplot(data.frame(gen = seq_along(sim_eq$L_vec), size = sim_eq$L_vec),
                 aes(x = gen, y = size)) +
  geom_line(colour = "steelblue") +
  theme_classic(base_size = 11) +
  labs(x = "Generation", y = "Array size", title = "Size trajectory")

p_load <- ggplot(data.frame(load = load), aes(x = load)) +
  geom_histogram(bins = 40, fill = "firebrick", alpha = 0.7, colour = "white") +
  theme_classic(base_size = 11) +
  labs(x = "Mismatches to consensus", y = "Count",
       title = "Mutation load distribution at equilibrium")

plot_grid(p_traj, p_load, ncol = 1)
```

## The fingerprint at equilibrium

What does the identical-pair structure look like when the array has been
evolving at equilibrium?

```{r fingerprint_eq, fig.height=8, cache=TRUE}
# Need to compute pairs for plotting
sim_eq_with_pairs <- sim_eq
sim_eq_with_pairs$ps <- pairs_identical(mono)

plot_ps_summary(sim_eq_with_pairs, title = "Equilibrium array (8000 gens)")
```

## Implications for ABC inference

The equilibrium model has an important implication for parameter inference:
**array size itself becomes an informative summary statistic.** Instead of
growing arrays to a preset target size, we can let the simulation run for a
fixed number of generations and see what size it reaches. The observed array
size then constrains the ratio of duplication to deletion rates.

This is more biologically realistic than the "grow to N" stopping rule,
because real repeat arrays in genomes reflect an evolutionary equilibrium,
not an arbitrary size cap.

Key parameters that determine the equilibrium:

- **Distal duplication rate** (per-array): Controls the constant input
- **Deletion rate** (per-unit): Controls size-dependent loss
- **Deletion chunk size**: Larger chunks = more material removed per event
- **Local duplication rate**: Adds to growth but doesn't create equilibrium alone

The mutation load at equilibrium is determined by `mu_total × n_generations`,
providing an independent constraint on evolutionary time.
