---
title: "Introduction to reCYCLing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to reCYCLing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4,
  out.width  = "100%"
)
set.seed(42)
library(data.table)
```

## Overview

**reCYCLing** simulates the evolution of tandem repeat arrays—sequences where
a short monomer unit is repeated many times in a row, as found in satellite DNA,
centromeric repeats, and other genomic repeat elements.

Starting from a small array of identical monomers, the simulator applies four
forces each generation:

| Process | Controls |
|---|---|
| **Local duplication** | Tandem copy–paste of a nearby chunk |
| **Distal duplication** | Copy–paste to a more distant position in the array |
| **Chunk deletion** | Removal of a block of consecutive monomers |
| **Per-base mutation** | Substitutions (and optionally indels) within each monomer |

---

## Running a simulation

```{r run_sim}
library(reCYCLing)

# Define a fixed ancestral monomer so results are reproducible
ancestor <- paste(sample(c("A","C","G","T"), 30, replace = TRUE),
                  collapse = "")

sim <- run_sim_ps(
  n                  = 1,         # one replicate
  init_l             = 30,        # monomer length (bp)
  init_k0            = 10,        # starting copy number
  init_sequence_type = "given",
  ancestor_seq       = ancestor,
  max_units          = 500,       # stop when array reaches 500 monomers
  mu_total           = 3e-5,      # per-base mutation rate
  p_local_dup        = 0.00015,   # local duplication probability per unit
  p_distal_dup       = 1.5e-6,    # distal duplication probability per unit
  p_del_chunk        = 0,         # no deletions in this example
  verbose            = FALSE
)
```

---

## Understanding the output object

`run_sim_ps()` returns a **named list with five elements**, each a list of
length `n` (one entry per replicate). For a single replicate (`i = 1`):

```{r output_structure}
names(sim)
```

### `monomers_list` — the final array

Each element is a `data.table` with one row per monomer in the final array.

```{r monomers}
mono <- sim$monomers_list[[1]]
head(mono)
nrow(mono)          # total monomers in the final array
```

Column descriptions:

| Column | Description |
|---|---|
| `num` | Position index along the array (1 = first monomer) |
| `bponly` | The monomer DNA sequence |
| `pos` | Sub-position within the locus (always 1 for simple arrays) |
| `chrom` | Chromosome label |
| `hap` | Haplotype label |
| `sample` | Sample label |
| `dir` | Strand direction |

The key columns for most analyses are **`num`** (position) and **`bponly`**
(sequence).

```{r monomers_summary}
# How many unique sequences are in the array?
length(unique(mono$bponly))

# Shannon entropy — higher = more diversity among monomers
shannon_entropy(mono$bponly)
```

### `ps_list` — all pairs of identical monomers

Each element is a `data.table` of every unordered pair of monomers that share
the same sequence. This is the foundation of the fingerprint plot.

```{r ps}
ps <- sim$ps_list[[1]]
head(ps[, c("bponly", "num1", "num2")])
nrow(ps)   # total number of identical pairs
```

The key columns are **`num1`** and **`num2`** (positions of the two identical
monomers) and **`bponly`** (their shared sequence).

### Trajectory vectors — how the array evolved over time

The three trajectory lists record statistics sampled **each generation**:

```{r trajectories}
# Array size (copy number) over generations
L_traj <- sim$L_vec_list[[1]]
cat("Generations simulated:", length(L_traj), "\n")
cat("Starting size:", L_traj[1], "| Final size:", tail(L_traj, 1), "\n")

# Shannon entropy over generations
H_traj <- sim$H_vec_list[[1]]
cat("Starting entropy:", round(H_traj[1], 3),
    "| Final entropy:", round(tail(H_traj, 1), 3), "\n")

# Unique monomer count over generations
N_traj <- sim$N_vec_list[[1]]
```

You can plot these trajectories to visualise how the array expanded and
diversified:

```{r traj_plot, fig.height=3}
library(ggplot2)

traj <- data.frame(
  gen    = seq_along(L_traj),
  size   = L_traj
)

ggplot(traj, aes(x = gen, y = size)) +
  geom_line() +
  theme_classic(base_size = 10) +
  xlab("Generation") +
  ylab("Array size (monomers)")
```

---

## Individual plot functions

Each diagnostic view of the array is available as its own function so you can
combine, customise, or inspect them independently.

### `plot_repeat_fingerprint()` — dot-plot of identical pairs

Each dot represents a pair of identical monomers, placed at their positions
along both axes. Structure near the diagonal reflects local duplications;
off-diagonal clusters reveal distal duplications.

```{r p_fingerprint, fig.height=5}
plot_repeat_fingerprint(sim, i = 1, ptsize = 0.5)
```

### `plot_pair_distances()` — distance between identical pairs

Plots the positional separation |num₁ − num₂| of every identical pair on a
log₁₀ scale. A peak near 1 indicates mostly tandem (local) duplications;
a heavy tail indicates distal duplications were important.

```{r p_dist}
plot_pair_distances(sim, i = 1)
```

### `plot_copy_numbers()` — copies per unique monomer sequence

For each distinct monomer sequence, counts how many copies exist in the array
and plots the distribution. High copy numbers mean many monomers share an
identical sequence—a signature of homogenisation by repeated duplication.

```{r p_cn}
plot_copy_numbers(sim, i = 1)
```

### `plot_mutation_load()` — mismatches relative to consensus

Computes the consensus sequence across all monomers and counts mismatches for
each individual monomer. Monomers with load = 0 are identical to the consensus;
higher loads indicate more diverged copies.

```{r p_ml}
plot_mutation_load(sim, i = 1)
```

### `plot_consensus_support()` — per-position consensus strength

At each alignment position, the proportion of monomers agreeing with the
consensus base is plotted as a histogram. Values near 1 mean that position
is strongly conserved; a flat distribution indicates heterogeneity.

```{r p_cs}
plot_consensus_support(sim, i = 1)
```

### `plot_ps_summary()` — all five panels combined

When you want a quick overview, `plot_ps_summary()` assembles all five plots
into a single figure using **cowplot**.

```{r summary, fig.height=6}
plot_ps_summary(sim, i = 1, title = "Single replicate summary")
```

---

## Sharing pre-computed counts across plots

`plot_mutation_load()` and `plot_consensus_support()` both require computing
the per-position counts with `counts_long_nogap()`. If you call both
functions, you can compute the counts once and pass them to each:

```{r counts_reuse}
mono   <- sim$monomers_list[[1]]
counts <- counts_long_nogap(mono$bponly)

p_ml <- plot_mutation_load(sim, i = 1, counts = counts)
p_cs <- plot_consensus_support(sim, i = 1, counts = counts)
```

---

## Comparing parameter settings

A common use-case is running several replicates under different mutation rates
or duplication probabilities and comparing the resulting entropy:

```{r compare, eval=FALSE}
sim_low  <- run_sim_ps(n = 3, max_units = 300, mu_total = 1e-5)
sim_high <- run_sim_ps(n = 3, max_units = 300, mu_total = 1e-4)

ent_low  <- get_sim_entropies(sim_low)
ent_high <- get_sim_entropies(sim_high)

ent_low$mu  <- "low"
ent_high$mu <- "high"

library(data.table)
ent_all <- rbind(ent_low, ent_high)

ggplot(ent_all, aes(x = mu, y = H)) +
  geom_jitter(width = 0.1) +
  theme_classic() +
  ylab("Shannon entropy (bits)")
```

---

## Function reference

| Function | Purpose |
|---|---|
| `run_sim_ps()` | Run `n` replicates of repeat-array evolution |
| `init_array()` | Initialise an array (used internally by `run_sim_ps`) |
| `pairs_identical()` | Find all pairs of identical monomers |
| `counts_long_nogap()` | Per-position base counts across aligned monomers |
| `mu_load_from_consensus()` | Mismatch count per monomer vs consensus |
| `shannon_entropy()` | Entropy (bits) of a monomer distribution |
| `shannon_entropyplus()` | Entropy plus diversity statistics |
| `get_sim_entropies()` | Entropy summary across all replicates |
| `kb_row_window_entropy()` | Sliding-window entropy along an array |
| `plot_repeat_fingerprint()` | Dot-plot of identical pairs |
| `plot_pair_distances()` | Distance distribution between identical pairs |
| `plot_copy_numbers()` | Copy-number distribution per unique monomer |
| `plot_mutation_load()` | Mutation load histogram |
| `plot_consensus_support()` | Consensus support histogram |
| `plot_ps_summary()` | Combined five-panel summary figure |
| `plot_knob_barcode()` | Linear barcode of a zoom window |
| `plot_circular_seqcounts()` | Circular base-composition plot |
