---
title: "Introduction to reCYCLing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to reCYCLing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%",
  dpi       = 300
)
set.seed(42)
library(data.table)
```

## Overview

**reCYCLing** simulates the evolution of tandem repeat arrays—sequences where
a short monomer unit is repeated many times in a row, as found in satellite DNA,
centromeric repeats, and other genomic repeat elements.

Starting from a small array of identical monomers, the simulator applies four
forces each generation:

| Process | Controls |
|---|---|
| **Local duplication** | Tandem copy–paste of a nearby chunk |
| **Distal duplication** | Copy–paste to a more distant position in the array |
| **Chunk deletion** | Removal of a block of consecutive monomers |
| **Per-base mutation** | Substitutions (and optionally indels) within each monomer |

---

## Running a simulation

```{r run_sim}
library(reCYCLing)

# Define a fixed ancestral monomer so results are reproducible
ancestor <- paste(sample(c("A","C","G","T"), 178, replace = TRUE),
                  collapse = "")

sim <- run_sim_ps(
  init_l             = 178,       # monomer length (bp)
  init_k0            = 10,        # starting copy number
  init_sequence_type = "given",
  ancestor_seq       = ancestor,
  max_units          = 10000,     # stop when array reaches 10 000 monomers
  mu_total           = 5e-5,      # per-base mutation rate
  p_local_dup        = 0.0004,    # local duplication probability per unit
  p_distal_dup       = 1e-5,      # distal duplication probability per unit
  p_del_chunk        = 1e-5,      # deletion probability per unit
  verbose            = FALSE
)
```

---

## Understanding the output

`run_sim_ps()` returns a named list with five elements:

```{r output_structure}
names(sim)
```

### `monomers` — the final array

A `data.table` with one row per monomer in the final array.

```{r monomers}
head(sim$monomers)
nrow(sim$monomers)   # total monomers in the final array
```

| Column | Description |
|---|---|
| `num` | Position index along the array (1 = first monomer) |
| `bponly` | The monomer DNA sequence |
| `dir` | Strand direction (`+` or `-`; reflects inversions from distal duplication) |

```{r monomers_summary}
# How many unique sequences are in the array?
length(unique(sim$monomers$bponly))

# Shannon entropy — higher = more diversity among monomers
shannon_entropy(sim$monomers$bponly)
```

### `ps` — all pairs of identical monomers

A `data.table` of every unordered pair of monomers that share the same
sequence. This is the foundation of the fingerprint plot.

```{r ps}
head(sim$ps[, c("bponly", "num1", "num2")])
nrow(sim$ps)   # total number of identical pairs
```

### Trajectory vectors — how the array evolved over time

```{r trajectories}
cat("Generations simulated:", length(sim$L_vec), "\n")
cat("Starting size:", sim$L_vec[1], "| Final size:", tail(sim$L_vec, 1), "\n")
```

```{r traj_plot, fig.height=3}
library(ggplot2)

traj <- data.frame(gen = seq_along(sim$L_vec), size = sim$L_vec)

ggplot(traj, aes(x = gen, y = size)) +
  geom_line() +
  theme_classic(base_size = 10) +
  xlab("Generation") +
  ylab("Array size (monomers)")
```

---

## Visualising results

Each diagnostic view of the array is available as its own plotting function.

### Repeat fingerprint — dot-plot of identical pairs

Each dot represents a pair of identical monomers, placed at their positions
along both axes. Diagonal streaks near the main diagonal reflect local (tandem)
duplications; off-diagonal clusters reveal distal duplications.

```{r p_fingerprint, fig.height=6}
plot_repeat_fingerprint(sim, ptsize = 0.01)
```

### Zooming in

The full fingerprint compresses 10,000 positions into a single figure. Use the
`zoom` parameter to inspect a sub-region in detail:

```{r p_fingerprint_zoom, fig.height=6}
plot_repeat_fingerprint(sim, zoom = c(4000, 5000), ptsize = 0.3,
                        gridalpha = 0.3)
```

### Pair distances — separation between identical pairs

The positional separation |num1 − num2| of every identical pair on a log10
scale. A peak near 1 indicates mostly tandem (local) duplications; a heavy tail
or second mode at larger distances indicates distal duplications were important.

```{r p_dist}
plot_pair_distances(sim)
```

### Copy numbers — copies per unique monomer sequence

For each distinct monomer sequence, counts how many copies exist in the array.
High copy numbers mean many monomers share an identical sequence—a signature of
homogenisation by repeated duplication.

```{r p_cn}
plot_copy_numbers(sim)
```

### Mutation load — mismatches relative to consensus

Computes the consensus sequence across all monomers and counts mismatches for
each individual monomer. Monomers with load = 0 are identical to the consensus;
higher loads indicate more diverged copies.

```{r p_ml}
plot_mutation_load(sim)
```

### Consensus support — per-position conservation

At each alignment position, the proportion of monomers agreeing with the
consensus base is plotted. Values near 1 mean that position is strongly
conserved; a spread distribution indicates heterogeneity.

```{r p_cs}
plot_consensus_support(sim)
```

### Combined summary — all five panels at once

`plot_ps_summary()` assembles all five plots into a single figure using
**cowplot**.

```{r summary, fig.height=10}
plot_ps_summary(sim, title = "Single replicate summary")
```

---

## Sharing pre-computed counts across plots

`plot_mutation_load()` and `plot_consensus_support()` both compute per-position
counts internally. If you call both, compute once and pass:

```{r counts_reuse}
counts <- counts_long_nogap(sim$monomers$bponly)

p_ml <- plot_mutation_load(sim, counts = counts)
p_cs <- plot_consensus_support(sim, counts = counts)
```

---

## Comparing parameter settings

A common use-case is running several simulations under different parameters and
comparing the resulting entropy:

```{r compare, eval=FALSE}
sims <- lapply(c(1e-5, 1e-4), function(mu) {
  run_sim_ps(max_units = 300, mu_total = mu)
})

ent <- rbindlist(lapply(seq_along(sims), function(i) {
  e <- get_sim_entropies(sims[[i]])
  e$mu <- c("low", "high")[i]
  e
}))

ggplot(ent, aes(x = mu, y = H)) +
  geom_point(size = 3) +
  theme_classic() +
  ylab("Shannon entropy (bits)")
```

---

## Function reference

| Function | Purpose |
|---|---|
| `run_sim_ps()` | Run a single repeat-array evolution simulation |
| `init_array()` | Initialise an array (used internally by `run_sim_ps`) |
| `pairs_identical()` | Find all pairs of identical monomers |
| `counts_long_nogap()` | Per-position base counts across aligned monomers |
| `mu_load_from_consensus()` | Mismatch count per monomer vs consensus |
| `shannon_entropy()` | Entropy (bits) of a monomer distribution |
| `shannon_entropyplus()` | Entropy plus diversity statistics |
| `get_sim_entropies()` | Entropy summary for a simulation result |
| `kb_row_window_entropy()` | Sliding-window entropy along an array |
| `plot_repeat_fingerprint()` | Dot-plot of identical pairs |
| `plot_pair_distances()` | Distance distribution between identical pairs |
| `plot_copy_numbers()` | Copy-number distribution per unique monomer |
| `plot_mutation_load()` | Mutation load histogram |
| `plot_consensus_support()` | Consensus support histogram |
| `plot_ps_summary()` | Combined five-panel summary figure |
| `plot_repeat_barcode()` | Linear barcode of a zoom window |
| `plot_circular_seqcounts()` | Circular base-composition plot |
